One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2019-11-22 paper.tex
slight tweaks to old draft

\begin{abstract}
Distance sampling methods aim to estimate the size and distribution of animal populations.  Traditionally this has been achieved using a two-stage approach by using first-stage estimates of detectability as an offset in a second-stage generalised additive model.  One-stage approaches have typically been implemented using MCMC and data augmentation methods. Here we present a new approach to one-stage distance sampling models, using iterated fits of latent Gaussian models fitted using integrated nested Laplace approximations (INLA) to simultaneously estimate the detectability and spatial distribution.  We model the data as a thinned log-Gaussian Cox process using computationally efficient Gaussian Markov random field approximations to Gaussian random fields.  We demonstrate the approach by applying the methods to point count survey data of Hawaiian \akepa{}. Our Bayesian one-stage approach allows sampling from the joint posterior of all model parameters and so the resulting summaries and plots naturally incorporate uncertainty in the observation model estimates.    
\end{abstract}

@@

The estimation of the size of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}.  Distance sampling methods aim to achieve this by using a spatially explicit sampling design and assumed detection model to estimate the detectability of animals as a function of distance from observer (in the case of point transects) or perpendicular distance to a transect line (for line transects) \citep{buckland_advanced_2008, buckland_distance_2015}.  Standard distance sampling approaches use a hybrid of design- and model-based inference to estimate population size.  The probability of detection is modelled as a function of distance and a randomized sampling design allows the construction of Horvitz-Thompson-like estimators of animal density \citep{horvitz_generalization_1952,  buckland_advanced_2008}.

@@

Our case study is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.  However, the point process viewpoint is not new and has been applied numerous times to line transect data \citep{buckland_model-based_2016, johnson_model-based_2010, hedley_spatial_2004, stoyan_remark_1982, hogmander_random_1991}.  We also further develop the point process viewpoint to allow point process models to be applied to data with incomplete location information.  It is common for distance sampling surveys to only record the distance to the observer and the observer location.  We show how point process methods can handle this type of data even though the exact location of the point is not known.

@@

Standard distance sampling approaches specify $g(s)$ as a function that decays with increasing distance.  The relevant distance depends on the survey type.  For line transects the perpendicular distance to the transect line is used.  For point transects the distance to the observer is used.  For the remainder of the paper we assume a point transect survey design.

@@

The thinned log-Gaussian Cox process likelihood with observed points at locations $\bm{Y} = (s_1, \ldots, s_n)^\intercal$ is then
\begin{equation}
\label{lgcp-likelihood}
\pi(\bm{Y} | \theta, \phi) = \exp\left(-\int\displaylimits_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s \right)\prod_{i=1}^n \lambda(s_i)g(s_i)
\end{equation}
The integral component of the likelihood does not usually have analytical solutions.  Replacing the integral with a weighted sum allows the log-likelihood to be rewritten as a weighted Poisson log-likelihood, which we describe below, adapting the approach of \cite{simpson_going_2016} for a fully observed point pattern.

@@

To simplify notation below we let $\tilde{\alpha}_{k} = (\alpha_{k1}, \ldots, \alpha_{kM})^\intercal$ and $\tilde{\alpha} = (\alpha_1^\intercal, \ldots, \alpha_K^\intercal)^\intercal$.  Similarly let $\tl_k = (\tl(s_{k1}), \ldots, \tl(s_{kM}))^\intercal$, $\tl_{int} = (\tl_1^\intercal, \ldots, \tl_K^\intercal)^\intercal$ and denote the intensity evaluated at the observed locations as $\tl_{obs} = (\tl(s_1), \ldots, \tl(s_n))^\intercal$

Then the approximate log-likelihood is

\begin{equation}
\label{approx-log-likelihood}
	\log \pi(\bm{Y}) \approx - \tilde{\alpha}^\intercal \tl_{int} + 1^\intercal\log\tl_{obs}
\end{equation}

This approximation can be written as a modified Poisson likelihood.  To see this let $\exp \eta = (\tl_{int}^\intercal, \tl_{obs}^\intercal)^\intercal$,
$\alpha = (\tilde{\alpha}^\intercal, 0_{n \times 1}^\intercal)^\intercal$ and construct a vector of pseudo-observations $y = (0_{KM\times 1}^\intercal, 1_{n \times 1}^\intercal)^\intercal$.  Then the approximate likelihood becomes
\begin{equation}
\pi(\bm{Y}) \approx \prod_{i=1}^{KM + n} \eta_i^{y_i}\exp(-\alpha_i\eta_i)
\end{equation}

This is similar to a Poisson likelihood with an offset term equal to $\log\alpha$ and can be specified using the Poisson likelihood implementation available in \texttt{R-INLA}.

@@

The log-intensity of animal locations is given by
\begin{equation*}
\log \lambda(s) = \beta_0 + \sum_v \beta_v z_v(s) + \xi(s)
\end{equation*}
where $\beta_0$ is an intercept parameter, the $z_v(s)$ are spatially indexed covariates with effect parameters $\beta_v$ and $\xi(s)$ is a zero-mean Gaussian random field with Mat\'ern covariance
\begin{equation}
C(s_1,s_2) = \frac{2^{1-\nu}}{4\pi\kappa^2\tau^2\Gamma(\nu)}(\kappa \|s_1-s_2\|)^{\nu}K_\nu(\kappa \|s_1-s_2\|)
\end{equation}
where \(\nu, \kappa, \tau\) are parameters and \(K_{\nu}\) is the modified Bessel function of the second kind.  All three parameters are not simultaneously identifiable \citep{zhang_inconsistent_2004} and it is conventional to assume a value for $\nu$ which controls the mean-square differentiability of the process.  We set $\nu = 1$ which is the default value in \texttt{R-INLA}. [[Would like to justify $\nu = 1$ somehow?]]

We use a Gaussian Markov random field approximation of $\xi(s)$ following the approach of \cite{lindgren_explicit_2011} based on specifying $\xi(s)$ via a stochastic partial differential equation.  Given a finite element mesh on the spatial domain with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$ we represent the GRF as $\xi(s) = \sum_l \xi_l \phi_l(s)$.  The parameters $\xi_1, \ldots, \xi_L$ form a Gaussian Markov random field with sparse prior precision matrix $\bm{Q}_{\xi} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices the elements of which are defined via an approximate solution to a stochastic partial differential equation. The precision parameter $\tau$ controls the marginal variance of the field and $\kappa$ is an inverse range parameter that controls the rate of decay of the covariance as distance between locations increases.  Further details of this Gaussian Markov random field approximation are given in the supplementary materials.

@@

The above specifies a model for all animal locations in the area of interest.  The log-intensity for observed points is $\log\tl(s) = \log\lambda(s) + \log g(s)$.  This presents a problem since $\log g(s)$ is not typically linear in its parameters. However, for any single set of parameter values $g(s)$ can be viewed as acting like an offset. If we could specify $g(s)$ as known then we could proceed with an inference approach suitable for additive linear predictors.

@@

% RJC 09/08/2019 Added case study text between % addtextstart and % addtextend markers.
% addtextstart
In this section, we apply the above methods to point transect data on the \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}), an internationally and federally endangered Hawaiian honeycreeper (\citealp{usfws_akepa_1970, birdlife_akepa_2016}) that is endemic to \hawaii{} Island, USA.  Large-scale, quantitative surveys of Hawaiian forest birds and their habitat commenced in the mid-1970s through the Hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986}. Information from the HFBS was used to update the listing and delisting of endangered species, and establish preserves that coincided with native bird hotspots, including Hakalau Forest National Wildlife Refuge that was the first wildlife refuge with the primary purpose to protect, conserve and manage native forests for threatened and endangered bird and plant species. Due to the broad-scale coverage and robust design HFBS has become the baseline to determine changes in bird species distributions, population sizes and trends in density patterns over time.

During the 20th century \akepa{} declined dramatically due to habitat modification \citep{scott_HFBS_1986, pratt_avifaunal_1994},  mosquito-transmitted avian diseases \citep{pratt_avifaunal_1994, atkinson_wildlife_1995}, introduced predators \citep{lepson_akepa_1997}, and food resources competitors \citep{lepson_akepa_1997}. \Akepa{} has a global abundance of 16,248 (95\%CI 10,074\textendash25,198) birds that has been restricted to five spatially distinct populations \citep{judge_akepa_2018}. Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau), supports the largest \akepa{} population that in 2012 was estimated at more than 11,000 birds \citep{camp_statespace_2016}. Maintaining and expanding the \akepa{} population at Hakalau is a primary conservation concern. Moreover, having unbiased and precise abundance estimates are required by land and resource managers for evaluating management actions and establishing management planning, and policy makers for decision-making processes.

@@

For the purposes of our analysis we selected a single survey from the \akepa{} time series that contained on broad sampling of the study area with sufficient numbers of detections to estimate detectability. In 2002, 195 points were sampled using point-transect distance sampling methods within the 3,061 ha open-forest study area of Hakalau (Fig. \ref{fig:2002studyareapointspt}). On 67 points 152 \akepa{} were detected. The number of detections within each sampling unit ranged from zero to 6.

@@

Second, we specify the thinning probability function model component.  The data contains only distances from the observer so we use equation \eqref{intensity-incomplete} to specify the log thinning probability function as
\begin{lstlisting}
thinning_function <- function(distance, lsig){
    log(2*pi) + log(distance) + log_halfnormal(distance, lsig)
    }
\end{lstlisting}
where \texttt{log\_halfnormal} is the log half-normal detection function $\log g(r | \sigma) = -1/2(r/\sigma)^2$ and the argument \texttt{lsig} is $\log(\sigma)$.  The dataset contains a column named \texttt{distance} that contains the observed distances.  The software will search for the appropriate columns using the names of the arguments specified in this function.  To allow the software to distinguish between named covariates and parameters - which both appear as arguments of the function to be estimated - we must also specify a model components object
\begin{lstlisting}
cmp <- ~ Intercept + grf(map = coordinates, model = matern) + lsig 
\end{lstlisting} 
This object names all model components to be estimated.  Note that the \texttt{grf} component is a name we have chosen and not a function.  The software allows user-specified names for the \texttt{f()} function known to users of \texttt{R-INLA}. Note that \texttt{coordinates} is a keyword that allows users to supply their data in the form of a spatial object using the \texttt{sp} package \citep{pebesma_spatial_2005}.

We can now specify the predictor formula using the components defined above.  
\begin{lstlisting}
fml <- coordinates + distance ~ Intercept + grf +
	thinning_function(distance, lsig)
\end{lstlisting}
To fit the model we specify starting values for the parameters, the domain of possible observed distances and fit the model using the \texttt{lgcp} function
\begin{lstlisting}
starting_values <- data.frame(grf = 0, lsig = 3.65, Intercept = 0)
W <- 60   # transect radius
distance_domain <- seq(.Machine$double.eps, W, length.out = 30)
lgcp(components = cmp, 
     data = obs,
     samplers = samplers,
     domain = list(distance = distance_domain),
     formula = fml,
     options = list(result = starting_values, max.iter = 30))
\end{lstlisting}
where our observed data is contained in a \texttt{SpatialPointsDataFrame} object named \texttt{obs}.  We also specify the \texttt{samplers} object, also a \texttt{SpatialPointsDataFrame}, which contains the locations of all sampling units.  Full scripts for the analysis are available in the supplementary materials.

@@

The most common method to analyse distance sampling data is a two-stage approach with a generalised additive model that includes a spatial smoothing spline.  The approach here of specifying a point process model and including an SPDE effect can seem fundamentally different.  However, these two approaches are related and an analysis of the \akepa{} data using the GAM approach provides similar point estimates.  There is a correspondence between SPDEs and smoothing splines: the smoothing spline is an estimator of the posterior mean of the SPDE \citep{kimeldorf_spline_2002}.  The SPDE and the smoothing spline play a similar role within the model and the resulting mean field map looks similar in both approaches.  
In addition, it can appear as though the point process model avoids the need to bin data into counts and so avoids some loss of information. This is not the case since we also applied the standard distance sampling assumption of constant intensity within each point transect.  This can be thought of as the point process version of binning data into counts since the location of points within transects does not inform the spatial distribution of points across the whole study region.  The main advantage of the log-Gaussian Cox process is that the discretisation of space occurs during the step of approximating the GRF.  This is in contrast to motivating the discretisation by the act of binning the data into counts so as to create a response variable that can be analysed using a GAM model.  We also note that \texttt{inlabru} provides functionality for fitting models with multiple likelihoods \citep{bachl_inlabru_2019}.  As such our approach is extendible to more complicated contexts such as marked point processes.

We believe the excursions approach is a valuable addition to the literature on communicating uncertainty in estimated surfaces.  For species distribution modelling this has the potential to be a useful tool for model evaluation and communication with stakeholders.  In particular the approach requires users to decide (i) a threshold or set of threshold values of interest and (ii) an acceptable probability of error. 
We believe that, in most situations, these decisions should be made by domain experts who have an understanding of the phenomena of interest, the set of possible decisions that may be taken on the basis of the analysis and the consequences of error.  Once these decisions have been made the statistician can then proceed accordingly, providing outputs  that inform on quantities of interest with a level of uncertainty that stakeholders have agreed is acceptable.  We note that the excursions approach applied here has the potential to be useful in other methods of species distribution mapping, environmental statistics and spatial statistics in general.  There is the potential to use excursions to motivate both the \textit{definition} and \textit{estimation} of concepts of interest to ecologists, such as a species' home range or the notion of occupancy.

>>>>

\begin{abstract}
Distance sampling methods aim to estimate the size of animal populations.  Incorporating spatial models of animal density has traditionally been achieved using a two-stage approach where first-stage estimates of detectability are used as an offset in a second-stage generalised additive model.  One-stage approaches have typically been implemented using MCMC and data augmentation methods. Here we present a new approach to one-stage distance sampling models, using iterated fits of latent Gaussian models fitted using integrated nested Laplace approximations (INLA) to simultaneously estimate the detectability and spatial distribution.  We model the data as a thinned log-Gaussian Cox process using computationally efficient Gaussian Markov random field approximations to Gaussian random fields.  We demonstrate the approach by applying the methods to point count survey data of a forest bird. Our Bayesian one-stage approach allows sampling from the joint posterior of all model parameters and so the resulting summaries and plots naturally incorporate uncertainty in the observation model estimates.    
\end{abstract}

@@

The estimation of the size of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}.  Distance sampling methods aim to achieve this by using a spatially explicit sampling design and assumed detection model to estimate the detectability of animals as a function of distance from observer (in the case of point transects) or perpendicular distance to a transect line (for line transects) \citep{buckland_advanced_2004, buckland_distance_2015}.  Standard distance sampling approaches use a hybrid of design- and model-based inference to estimate population size.  The probability of detection is modelled as a function of distance and a randomized sampling design allows the construction of Horvitz-Thompson-like estimators of animal density \citep{horvitz_generalization_1952,  buckland_advanced_2004}.

@@

Our case study is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.  However, the point process viewpoint is not new and has been applied numerous times to line transect data \citep{buckland_model-based_2016, johnson_model-based_2010, hedley_spatial_2004,  hogmander_random_1991, stoyan_remark_1982}.  We also further develop the point process viewpoint to allow point process models to be applied to data with incomplete location information.  It is common for distance sampling surveys to only record the distance to the observer and the observer location.  We show how point process methods can handle this type of data even though the exact location of the point is not known.

@@

Standard distance sampling approaches specify $g(s)$ as a function that decays with increasing distance.  The distance measured depends on the type of survey.  For line transects the perpendicular distance to the transect line is used.  For point transects the horizontal distance to the observer is the relevant distance.  For the remainder of the paper we assume a point transect survey design.

@@

The thinned log-Gaussian Cox process likelihood with observed points at locations $\bm{Y} = (s_1, \ldots, s_n)^\intercal$ is then
\begin{equation}
\label{lgcp-likelihood}
\pi(\bm{Y}) = \exp\left(-\int\displaylimits_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s \right)\prod_{i=1}^n \lambda(s_i)g(s_i)
\end{equation}
where we have suppressed parameter vector notation.  The integral component of the likelihood does not usually have analytical solutions.  Replacing the integral with a weighted sum allows the log-likelihood to be rewritten as a weighted Poisson log-likelihood, which we describe below, adapting the approach of \cite{simpson_going_2016} for a fully observed point pattern.

@@

To simplify notation below we let $\tilde{\alpha}_{k} = (\alpha_{k1}, \ldots, \alpha_{kM})^\intercal$ and $\tilde{\alpha} = (\alpha_1^\intercal, \ldots, \alpha_K^\intercal)^\intercal$.  Similarly let $\tl_k = (\tl(s_{k1}), \ldots, \tl(s_{kM}))^\intercal$, $\tl_{int} = (\tl_1^\intercal, \ldots, \tl_K^\intercal)^\intercal$ and denote the intensity evaluated at the observed locations as $\tl_{obs} = (\tl(s_1), \ldots, \tl(s_n))^\intercal$.  Then the approximate log-likelihood is
\begin{equation}
\label{approx-log-likelihood}
	\log \pi(\bm{Y}) \approx - \tilde{\alpha}^\intercal \tl_{int} + 1^\intercal\log\tl_{obs}
\end{equation}
This approximation can be written as a modified Poisson likelihood.  To see this let $\exp \eta = (\tl_{int}^\intercal, \tl_{obs}^\intercal)^\intercal$,
$\alpha = (\tilde{\alpha}^\intercal, 0_{n \times 1}^\intercal)^\intercal$ and construct a vector of pseudo-observations $y = (0_{KM\times 1}^\intercal, 1_{n \times 1}^\intercal)^\intercal$.  Then the approximate likelihood becomes
\begin{equation}
\pi(\bm{Y}) \approx \prod_{i=1}^{KM + n} \eta_i^{y_i}\exp(-\alpha_i\eta_i)
\end{equation}
This is similar to a Poisson likelihood with a $\log\alpha$ offset and can be specified using the Poisson likelihood implementation available in \texttt{R-INLA}.

@@

The log-intensity of animal locations is given by
\begin{equation*}
\log \lambda(s) = \beta_0 + \sum_v \beta_v z_v(s) + \xi(s)
\end{equation*}
where $\beta_0$ is an intercept parameter, the $z_v(s)$ are spatially indexed covariates with effect parameters $\beta_v$ and $\xi(s)$ is a zero-mean GRF with Mat\'ern covariance
\begin{equation}
C(s_1,s_2) = \frac{2^{1-\nu}}{4\pi\kappa^2\tau^2\Gamma(\nu)}(\kappa \|s_1-s_2\|)^{\nu}K_\nu(\kappa \|s_1-s_2\|)
\end{equation}
where \(\nu, \kappa, \tau\) are parameters and \(K_{\nu}\) is the modified Bessel function of the second kind.  All three parameters are not simultaneously identifiable \citep{zhang_inconsistent_2004} and it is conventional to assume a value for $\nu$ which controls the mean-square differentiability of the process.  We set $\nu = 1$ which is the default value in \texttt{R-INLA}. [[Would like to justify $\nu = 1$ somehow?]]

We use a GMRF approximation of $\xi(s)$ following the approach of \cite{lindgren_explicit_2011} based on specifying $\xi(s)$ via a SPDE.  Given a finite element mesh on the spatial domain with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$ we represent the GRF as $\xi(s) = \sum_l \xi_l \phi_l(s)$.  The parameters $\xi_1, \ldots, \xi_L$ form a GMRF with sparse prior precision matrix $\bm{Q}_{\xi} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices the elements of which are defined via an approximate solution to a SPDE. The parameter $\tau$ is a precision parameter and $\kappa$ an inverse range parameter that controls the rate of decay of the covariance as distance between locations increases.  Further details of the GMRF approximation are given in the supplementary materials.

@@

The log-intensity for observed points (the thinned intensity) is $\log\tl(s) = \log\lambda(s) + \log g(s)$.  This presents a problem since $\log g(s)$ is not typically linear in its parameters. However, for any single set of parameter values $g(s)$ can be viewed as an offset term. If we could specify $g(s)$ as known then we could proceed with an inference approach suitable for additive linear predictors.

@@

% RJC 09/08/2019 Added case study text between % addtextstart and % addtextend markers.
% addtextstart
In this section, we apply the above methods to point transect data on the \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}), an internationally and federally endangered Hawaiian honeycreeper (\citealp{usfws_akepa_1970, birdlife_akepa_2016}) that is endemic to \hawaii{} Island, USA.  Large-scale, quantitative surveys of Hawaiian forest birds and their habitat commenced in the mid-1970s through the Hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986}. Information from the HFBS was used to update the listing and delisting of endangered species, and establish preserves that coincided with native bird hotspots, including Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau) that was the first wildlife refuge with the primary purpose to protect, conserve and manage native forests for threatened and endangered bird and plant species. Due to the broad-scale coverage and robust design HFBS has become the baseline to determine changes in bird species distributions, population sizes and trends in density patterns over time.

During the 20th century \akepa{} declined dramatically due to habitat modification \citep{scott_HFBS_1986, pratt_avifaunal_1994},  mosquito-transmitted avian diseases \citep{pratt_avifaunal_1994, atkinson_wildlife_1995}, introduced predators \citep{lepson_akepa_1997}, and food resources competitors \citep{lepson_akepa_1997}. \Akepa{} has a global abundance of approximately 16,200 (95\%CI 10,000\textendash25,200) birds that has been restricted to five spatially distinct populations \citep{judge_akepa_2018}. Hakalau supports the largest \akepa{} population that in 2012 was estimated at more than 11,000 birds \citep{camp_statespace_2016}. Maintaining and expanding the \akepa{} population at Hakalau is a primary conservation concern. Moreover, having unbiased and precise abundance estimates are required by land and resource managers for evaluating management actions and establishing management planning, and policy makers for decision-making processes.

@@

For the purposes of our analysis we selected a single survey from the \akepa{} time series that contained on broad sampling of the study area with sufficient numbers of detections to estimate detectability. In 2002, 289 points were sampled using point-transect distance sampling methods within the 3,061 ha open-forest study area of Hakalau (Fig. \ref{fig:2002studyareapointspt}). 276 \akepa{} were detected on 121 point transects. The number of detections within each sampling unit ranged from zero to 6.

@@

Second, we specify the thinning probability function model component.  The data contains only distances from the observer so we use equation \eqref{intensity-incomplete} to specify the log thinning probability function as
\begin{lstlisting}
thinning_function <- function(distance, lsig){
    log(2*pi) + log(distance) + log_halfnormal(distance, lsig)
    }
\end{lstlisting}
where \texttt{log\_halfnormal} is the log half-normal detection function $\log g(r | \sigma) = -1/2(r/\sigma)^2$ and the argument \texttt{lsig} is $\log(\sigma)$.  The dataset contains a column named \texttt{distance} that contains the observed distances.  The software will search for the appropriate columns using the names of the arguments specified in this function.  To allow the software to distinguish between named covariates and parameters - which both appear as arguments of the function to be estimated - we must also specify a model components object
\begin{lstlisting}
cmp <- ~ Intercept + grf(map = coordinates, model = matern) + lsig 
\end{lstlisting} 
This object names all model components to be estimated.  Note that the \texttt{grf} component is a name we have chosen and not a function.  The software allows user-specified names for model components.  This feature replaces the \texttt{f()} function known to users of \texttt{R-INLA} and allows the user to keep track of multiple model components rather that calling \texttt{f()} repeatedly.  Note that \texttt{coordinates} is a keyword that allows users to supply their data in the form of a spatial object using the \texttt{sp} package \citep{pebesma_spatial_2005}.

We can now specify the predictor formula using the components defined above.  
\begin{lstlisting}
fml <- coordinates + distance ~ Intercept + grf +
	thinning_function(distance, lsig)
\end{lstlisting}
To fit the model we specify starting values for the parameters, the domain of possible observed distances and fit the model using the \texttt{lgcp} function
% UPDATE THE W WHEN THIS CHANGES
\begin{lstlisting}
starting_values <- data.frame(grf = 0, lsig = 3.65, Intercept = 0)
W <- 60   # transect radius
distance_domain <- seq(.Machine$double.eps, W, length.out = 30)
lgcp(components = cmp, 
     data = obs,
     samplers = samplers,
     domain = list(distance = distance_domain),
     formula = fml,
     options = list(result = starting_values, max.iter = 30))
\end{lstlisting}
where our observed data is contained in a \texttt{SpatialPointsDataFrame} object named \texttt{obs}.  We also specify the \texttt{samplers} object, also a \texttt{SpatialPointsDataFrame}, which contains the locations of all sampling units.  Full scripts for the analysis are available in the supplementary materials.

@@

The most common method to analyse combine spatial models and distance sampling data is a two-stage approach with a generalised additive model that includes a spatial smoothing spline.  The approach here of specifying a point process model and including an SPDE effect can seem fundamentally different.  However, these two approaches are related and an analysis of the \akepa{} data using the GAM approach provides similar point estimates.  There is a correspondence between SPDEs and smoothing splines: the smoothing spline is an estimator of the posterior mean of the SPDE \citep{kimeldorf_spline_2002}.  The SPDE and the smoothing spline play a similar role within the model and the resulting mean field map looks similar in both approaches.  
In addition, it can appear as though the point process model avoids the need to bin data into counts and so avoids some loss of information. This is not the case since we also applied the standard distance sampling assumption of constant intensity within each point transect.  This can be thought of as the point process version of binning data into counts since the location of points within transects does not inform the spatial distribution of points across the whole study region.  The main advantage of the log-Gaussian Cox process is that the discretisation of space occurs during the step of approximating the GRF.  This is in contrast to motivating the discretisation by the act of binning the data into counts so as to create a response variable that can be analysed using a GAM model.  We also note that \texttt{inlabru} provides functionality for fitting models with multiple likelihoods \citep{bachl_inlabru_2019}.  As such our approach is extendible to more complicated contexts such as marked point processes.

We believe the excursions approach is a valuable addition to the literature on communicating uncertainty in estimated surfaces.  For species distribution modelling this has the potential to be a useful tool for model evaluation and communication with stakeholders.  In particular the approach requires users to \textit{a priori} decide on (i) a threshold or set of threshold values of interest and (ii) an acceptable probability of error. 
We believe that, in most situations, these decisions should be made by domain experts who have an understanding of the phenomena of interest, the set of possible decisions that may be taken on the basis of the analysis and the consequences of error.  Once these decisions have been made the statistician can then proceed accordingly, providing outputs  that inform on quantities of interest with a level of uncertainty that stakeholders have agreed is acceptable.  We note that the excursions approach applied here has the potential to be useful in other methods of species distribution mapping, environmental statistics and spatial statistics in general.  There is the potential to use excursions to motivate both the \textit{definition} and \textit{estimation} of concepts of interest to ecologists, such as a species' home range or the notion of occupancy.
